Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-05-18 Manuscript/main.tex
Add Flowchart to the table

\received{2017-01-01}

\revised{2017-02-01}

\accepted{2017-03-01}

@@

\title{A demonstration of the \LaTeX class file for Statistics in
Medicine with Rmarkdown}

@@

\abstract{TODO: Change this accordingly; Lorem ipsum dolor sit amet,
consectetur adipiscing elit. Aenean ut elit odio. Donec fermentum tellus
neque, vitae fringilla orci pretium vitae. Fusce maximus finibus
facilisis. Donec ut ullamcorper turpis. Donec ut porta ipsum. Nullam
cursus mauris a sapien ornare pulvinar. Aenean malesuada molestie erat
quis mattis. Praesent scelerisque posuere faucibus. Praesent nunc nulla,
ullamcorper ut ullamcorper sed, molestie ut est. Donec consequat libero
nisi, non semper velit vulputate et. Quisque eleifend tincidunt ligula,
bibendum finibus massa cursus eget. Curabitur aliquet vehicula quam non
pulvinar. Aliquam facilisis tortor nec purus finibus, sit amet elementum
eros sodales. Ut porta porttitor vestibulum. Integer molestie, leo ut
maximus aliquam, velit dui iaculis nibh, eget hendrerit purus risus sit
amet dolor. Sed sed tincidunt ex. Curabitur imperdiet egestas tellus in
iaculis. Maecenas ante neque, pretium vel nisl at, lobortis lacinia
neque. In gravida elit vel volutpat imperdiet. Sed ut nulla arcu. Proin
blandit interdum ex sit amet laoreet. Phasellus efficitur, sem hendrerit
mattis dapibus, nunc tellus ornare nisi, nec eleifend enim nibh ac
ipsum. Aenean tincidunt nisl sit amet facilisis faucibus. Donec odio
erat, bibendum eu imperdiet sed, gravida luctus turpis.}

\keywords{Class file; \LaTeX; Statist. Med.; Rmarkdown;}

@@

  \begin{itemize}
  \tightlist
  \item
    What are the methods
  \item
    What are their problems
  \end{itemize}
\item
  Introduce the wiggliness matrix
\item
  Our method
\item
  The implication of our method
\item
  Structure of the rest of the paper
\end{itemize}

Nonlinear effect modeling has long history. Early work of GAMs can date
back to 1979 when \citet{Cleveland1979} proposed local smoothing for
bivariate analysis. Many mathematical and statistical models has been
proposed since, see xxx{[}\^{} TODO: insert an review citation{]}.

Many machine learning algorithms can also model nonlinear effect.
However, a wide range of machine learning models are confined by the
``black box'' nature, rendering little help on understanding the
etiology of diseases. Semi-parametric and non-parametric regression
models, are more popular thanks to its interpretability and flexibility.
Among those, generalized additive model (GAM) proposed in the seminal
work of \citet{Hastie1987} gains the most popularity thanks to its
flexibility and easiness to incorporate many different smoothing
functions. The backfitting algorithm estimates each smooth function of
an additive model iteratively using partial residuals. The backfitting
algorithm is compatible with almost any smoothing or modeling technique
to be used as the nonparametric components. However, it is unclear how
to estimate the degree of smoothness when necessary, for example for
smoothing splines. The computation cost for cross-validating the smooth
parameters can be expensive. \citet[PP.318-320]{Wood2017} provides more
discussion on the problem.

Among all the choices of smoothing function, one particular family stood
out, splines. Splines are piece-wise polynomial functions to interpolate
nonlinear signals. Most commonly used spline is cubic spline, thanks to
its smoothness property. \citep[PP.197]{Wood2017} Early spline research
focused on regression splines, where number and location of knots are
pre-determined and ordinary least squares is used to estimate the smooth
function. However, the specification of knots is not a trivial task, as
the the positioning of knots are influential on the performance of the
model. As an improvement, researchers proposed smoothing splines where
as many knots are used and later shrunk using penalized algorithms. The
smoothing spline concept was first introduced
\citep{OSullivan1986, OSullivan1988}, and later generalized and made
famous by \citet{Eilers1996}, called P-spline. Independently,
\citet{Ruppert1999} proposed similar idea. Later, Bayesian GAMs were
introduced as an alternative to address measurement error in covariates
\citep{Berry2002} or local adaptive smoothing \citep{Lang2004}.

Generalized additive model (GAM) was first proposed by the seminal work:
a response variable, \(Y\), given a \(p\)-dimensional vector of
covariates \(\boldsymbol{x}\) can be expressed as the summation of
smooth functions of \(\boldsymbol{x}\), written as \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),\qquad E[f_j(x_j)] = 0.
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. Hence, GAM solve the model fitting problem for many
distributional outcomes, such as Poisson, binomial, gamma and normal
models. Since the appearance of GAMs, it has been applied in many area
of research, such as xxx, xxx .

In many area of research, high dimensional analysis becomes predominant,
for example sequencing data analysis, electric health record data
analysis. Previous analysis heavily relies on linear models. Recently,
there are more focus putting on modeling the non-linear relationships in
high dimensional data analysis. Hence, developing more technically
matured models is imminent. Nevertheless expanding the GAM framework for
high-dimensional applications is not a trivial task. The difficulties
mainly come in two-fold: grouping nature of the basis functions, and the
balance between spline smoothness and variable sparsity. For GAM models,
a smoothing function is consists of multiple basis functions. The basis
functions of a variable is inherently grouped. The grouping structure
should be properly taken care of for accurate estimation. Moreover,
shrinkage is previously used solely for sparsity of the model in high
dimensional data analysis, or smoothness in GAM models. How to
organically combine the two together remains as a troubling question.

\hypertarget{spike-and-slab-prior}{%
\paragraph{Spike-and-slab prior}\label{spike-and-slab-prior}}

The term ``spike-and-slab'' was first coined in \citet{Mitchell1988}
describing the shape of the prior. It is a mixture distribution,
comprised with a skinny spike density \(f_{spike}(\cdot)\) for weak
signals and a flat slab density \(f_{slab}(\cdot)\) for large signals,
mathematically \[
 \beta_i|\gamma_i \sim (1-\gamma_i)f_{spike}(\beta_i) + \gamma_if_{slab}(\beta_i).
\]

@@

In this dissertation, we aim to address the challenges of non-linear
signal modeling in high-dimensional application using Bayesian
Hierarchical model. Specifically, we will propose a new Bayesian
spike-and-slab prior for the generalized additive models and Cox spline
models. The contribution of the proposed prior comes in two-fold: 1) it
supports simultaneous variable selection and function selection; 2) it
provides robust estimation and prediction for the coefficients and
outcomes. As previously described, simultaneous variable selection and
its functional form selection is not a trivial problem. Nevertheless,
the Bayesian spike-and-slab prior is a perfect solution to the problem,
as the latent binary indicator can be used to for variable selection and
the spike and slab densities are used to adaptive adjust the smoothness
of the spline function. Moreover, compared to currently available
Bayesian spline models which mostly conduct a `all-in-all-out' strategy,
i.e.~either including a non-linear form of a variable or excluding the
variable completely, we offer a more flexible solution, rending one of
three possibilities: no effect of a predictor, only linear effect of a
predictor, or non-linear effect. This simplifies the process of
determining what effect a predictor have on the outcome, without
involving hypothesis testing where the degree of freedom for non-linear
effect is complicated to estimate due to the shrinkage. Moreover, out
model inherent the the robust estimation from Bayesian models.

@@

@@

\begin{verbatim}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
\end{verbatim}

@@

\begin{verbatim}
if (!require("xtable")) install.packages("xtable")
\end{verbatim}

\begin{verbatim}
## Loading required package: xtable
\end{verbatim}

\begin{verbatim}
xt <- xtable(head(cars), caption = "A table", label = "tab:table")
print(xt, comment = FALSE)
\end{verbatim}

>>>>

\received{2021-01-01}

\revised{2021-02-01}

\accepted{2021-03-01}

@@

\title{Spike-and-Slab Generalized Additive Models and Fast Algorithms
for High-Dimensional Data}

@@

\abstract{Spike-and-slab priors have been extensively studied for
variable selection, and recently been extended to function selection in
generalized additive models (GAMs). Most of the previous works on
fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC)
algorithms to identify the promising subsets of functions. The
computational burden of MCMC makes these models less feasible,
especially in high-dimensional setting. Besides, most of the current
spike-and-slab GAM consider an ``all-in-all-out'' approach for function
selection, rendering less flexibility to answer if non-linear effects
are necessary in the model. We propose Bayesian hierarchical generalized
additive models for function selection and prediction. The proposed
models employ spike-and-slab priors, mixture normal and mixture double
exponential, for smooth and sparse solution. Two EM based algorithms,
EM-IWLS and EM-Coordinate Descent, are developed for accelerated
computation. Simulation studies and real data analysis demonstrated
improved performance against the state-of-art models, mgcv and COSSO.
The software implementation of the proposed methods is freely available
via the R package BHAM.}

\keywords{Spike-and-Slab Priors; High-Dimensional Data; Generalized
Additive Models; EM-IWLS; EM-Coordinate Decent}

@@

  \begin{itemize}
  \tightlist
  \item
    What are the methods
  \item
    What are their problems
  \end{itemize}
\item
  Introduce the wiggliness matrix
\item
  Our method
\item
  The implication of our method
\item
  Fast
\item
  Structure of the rest of the paper
\end{itemize}

Nonlinear effect modeling has long history. Early work of GAMs can date
back to 1979 when \citet{Cleveland1979} proposed local smoothing for
bivariate analysis. Since then, many mathematical and statistical models
has been proposed, see xxx\footnote{TODO: insert an review citation}.
With recent advancement of computing technology, many machine learning
algorithms that is computationally intensive start to gainmuch
popularity in outcome prediction when non-linear effects assumed. Such
models generally have improved prediction accuracy over classic
statistical models, but have a fatal flaw, lack of transparency and
interpretability. The primary objective of biomedical studies is to
understand the etiology of diseases, which requires not only the
knowledge if a biomarker is predictive of the disease, but also how the
biomarker cause the disease. ``Black box'' models could less likely to
be helpful.

As a balance between interpretability and model flexibility,
semi-parametric and non-parametric regression models are studied. Among
those, generalized additive model (GAM) proposed in the seminal work of
\citet{Hastie1987} gains the most popularity thanks to its flexibility
and easiness to incorporate many different smoothing functions. A
response variable, \(Y\), that is assumed to follow some exponential
family distribution \(EF(\mu. \phi)\), can be expressed as the summation
of smooth functions of a given \(p\)-dimensional vector of covariates
\(\boldsymbol{x}\), written as \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),\qquad E[f_j(x_j)] = 0.
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. GAM solves the model fitting problem for many distributional
outcomes, such as Poisson, binomial, gamma and normal models. In the
same article, @\citet{Hastie1987} proposed the backfitting algorithm
that estimates each smooth function of an additive model iteratively
using partial residuals. The backfitting algorithm is compatible with
almost any smoothing or modeling technique to be used as the
nonparametric components. However, it is unclear how to estimate the
degree of smoothness when necessary, for example for smoothing splines.
The computation cost for cross-validating the smooth parameters can be
expensive. \citet[PP.318-320]{Wood2017} provides more discussion on the
problem.

In many area of modern biomedical research, dealing with
high-dimensional data, also known as \(p >> n\) problem, is common, for
example sequencing data analysis, electric health record data analysis.
Many statistical models have been proposed for such problem, for
example, penalized models and Bayesian hierarchical models. However, up
to this point, most efforts of methodology development focus on modeling
linear effect in high dimensional setting. It is still not clear what is
the best practice on how to model non-linear effects in high-dimensional
setting. Similar to expanding classic models to high-dimensional models,
expanding the GAM framework for high-dimensional applications is not a
trivial task. The difficulties mainly come in two-fold: grouping nature
of the basis functions, and the balance between spline smoothness and
variable sparsity. For GAM models, a smoothing function is consists of
multiple basis functions. The basis functions of a variable is
inherently related. The grouping structure should be properly taken care
of for accurate estimation. Moreover, shrinkage is previously used
solely for sparsity of the model in high dimensional regression models,
or smoothness in GAMs. How to organically combine the two together
remains as a troubling question.

For high-dimensional data analysis, a common Bayesian modeling approach
is to impose a particular family of priors, the spike-and-slab (SS)
priors on the coeffcient in a regression modelss. The term
``spike-and-slab'' was first coined in \citet{Mitchell1988} describing
the shape of the prior. It is a mixture distribution, comprised with a
skinny spike density \(f_{spike}(\cdot)\) for weak signals and a flat
slab density \(f_{slab}(\cdot)\) for large signals, mathematically \[
 \beta_i|\gamma_i \sim (1-\gamma_i)f_{spike}(\beta_i) + \gamma_if_{slab}(\beta_i).
\]

@@

In this article, we aim to address the challenges of non-linear signal
modeling in high-dimensional application using Bayesian Hierarchical
model. Specifically, we will propose a new Bayesian spike-and-slab prior
for the generalized additive models and Cox spline models. The
contribution of the proposed prior comes in two-fold: 1) it supports
simultaneous variable selection and function selection; 2) it provides
robust estimation and prediction for the coefficients and outcomes. As
previously described, simultaneous variable selection and its functional
form selection is not a trivial problem. Nevertheless, the Bayesian
spike-and-slab prior is a perfect solution to the problem, as the latent
binary indicator can be used to for variable selection and the spike and
slab densities are used to adaptive adjust the smoothness of the spline
function. Moreover, compared to currently available Bayesian spline
models which mostly conduct a `all-in-all-out' strategy, i.e.~either
including a non-linear form of a variable or excluding the variable
completely, we offer a more flexible solution, rending one of three
possibilities: no effect of a predictor, only linear effect of a
predictor, or non-linear effect. This simplifies the process of
determining what effect a predictor have on the outcome, without
involving hypothesis testing where the degree of freedom for non-linear
effect is complicated to estimate due to the shrinkage. Moreover, out
model inherent the the robust estimation from Bayesian models.

@@

\includegraphics[width=24.76in]{Fig/SS_prior_flow_chart}

@@

@@
